---
title: "Tutorial 3"
author: "Laura Zywietz Rolon"
date: "11/3/2019"
output: html_document
header-includes :
- \usepackage{enumerate}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Inverse Transform Method

**Task 1**

Use the *generalized inverse transform method* to sample $X \sim f_{X}$
\[ f_{X} = \frac{1}{\pi} \frac{1}{\sqrt{x(1-x)}}, \qquad \text{for} \quad 0 < x < 1\]

a) derive the c.d.f. $F_{X}$ and its inverse
b) confirm numerically that your derivation is correct by computing $F_{X}^{-1}(F_{X}(x))$
c) use the inverse function method to sample $X \sim f_{X}$, plot a histogram to confirm visually that the distribution lines up with the target p.d.f.

**Solution 1**

a) c.d.f.: \[ F_{X}(x) = \int_{0}^{x} \frac{1}{\pi} \frac{1}{\sqrt{x'(1-x')}}dx' \stackrel{x' = y^2}{=} \frac{1}{\pi} \int_{0}^{\sqrt{x}} \frac{1}{\sqrt{y^2(1-y^2)}}2ydy = \frac{2}{\pi} \int_{0}^{\sqrt{x}} \frac{1}{\sqrt{1-y^2}}dy = \frac{2}{\pi} \lbrack \arcsin(\sqrt{x})\rbrack_{0}^{\sqrt{x}} = \frac{2}{\pi} \arcsin(\sqrt{x}) \]  
Inverse of $F_{X}$: \[F^{-1}(y) = \sin^2(\dfrac{\pi y}{2})\]

b)  \[F_{X}^{-1}(F_{X}(x)) = \sin^2(\dfrac{\pi F_{X}(x) }{2}) = \sin^2(\dfrac{\frac{2}{\pi} \arcsin(\sqrt{x})\pi}{2}) = x \]

c)
```{r}
#Inverse Transform Method
N = 10000
u = runif(N)

f = function(y){
  sin(pi/2*y)*sin(pi/2*y)
}

inversef = function(x){
  2/pi * asin(sqrt(x))
}

d = function(x){
  1/pi/sqrt(x*(1-x))
}

#n = c(1:100)/100
#plot(n, inversef(f(n)))

hist(f(u), main = "", breaks= 40, freq=FALSE, xlab = "x", ylab= "fX")
points(u, d(u), col= "blue")
#points(c(1:100)/100, d(c(1:100)/100))
```
**Task 2**

Write a computer program which implements a linear congruential random number generator given by the following recurrence relation:
\[ X_{i+1} = a \cdot X_{i} \quad \text{mod} \quad m\]
and study its properties for different pairs of
\[ (a, m) = (7^5, 2^{31}-1), (19, 6788793), (65539, 2^{31})\]
i.e. calculate its mean value and variance. Do they agree with your expectations?
The last (a,m) pair corresponds to an infamous random number generator named RANDU.

Finally, generate plots of order 10000 points of
\[ (X_{i}/m, X_{i+1}/m, X_{i+2}/m), \qquad i = 1,2,...\]
(3d scatter plots). Can you visually observe correlations?
You will see why RANDU became infamous.

**Solution 2**

```{r}
#Linear Congruential Random Number Generator
lCRNG <- function(a, m, v){
  (a * v) %% m
}

generateRN <- function(a, m, N){
  v <- c(1:N)
  v[1] <- 42
  for(i in c(2:N)){
    v[i] <- lCRNG(a, m, v[i-1])
  }
  return(v)
}

#Execution
a <- c(7**5, 19, 65539)
m <- c(2**31-1, 6788793, 2**21)
N <- 10000
r <- matrix(nrow = N, ncol = 3)
meanValue <- c(1:3)
varianceValue <- c(1:3)
for(i in c(1:3)){
  r[ ,i] <- generateRN(a[i], m[i], N)/m[i]
  meanValue[i] <- mean(r[ ,i])
  varianceValue[i] <- var(r[, i])
}
#currentVal <- 1
#r <- generateRN(a[currentVal], m[currentVal], N)
#print(r)
hist(r[ ,1], main="(a, m) = (7**5, 2**31-1)", freq=FALSE)
hist(r[ ,2], main="(a, m) = (19, 6788793)", freq=FALSE)
hist(r[ ,3], main="(a, m) = (65539, 2**21)", freq=FALSE)

# print("Mean:")
# print(meanValue)
# print("Variance:")
# print(varianceValue)

#scatterplot
library(scatterplot3d)
library(rgl)

idx <- c(1:9998)
scatterplot3d(x=r[idx, 3], y=r[idx+1, 3], z=r[idx+2, 3], angle=20, xlab="x", ylab="y", zlab="z")
#plot3d(x=r[idx, 3], y=r[idx+1, 3], z=r[idx+2, 3], angle=20, xlab="x", ylab="y", zlab="z")

```
