---
title: "Tutorial8"
author: "Laura Zywietz Rolon"
date: "12/4/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulated Annealing

Consider the following data:

$$
\begin{table}
  \centering
  \begin{tabular}{c|c|c}
  t & y & dy\\
  \hline
  1 & 0.94 & 0.11\\
  2 & 1.95 & 0.10\\
  3 & 3.15 & 0.13\\
  \end{tabular}
\end{table}
$$

to which we want to fit the two-parameter function 
\[f(t;a,b) = at +b\]
minimizing the $\chi^2$. We assume that the data points are independent.
Implement simulated annealing to minimize the $\chi^2$ function. Compare your result with the standard estimator for linear regressions.
For your implementation

* make sure you keep track of the minimal value for the $\chi^2$-function and the corresponding parameters the algorithm found.
* optimise the proposal distribution, which is typically a Gaussian distribution with a certain width.
* optimise the number of iterations you perform at fixed temperature $T$.
* optimise your strategy of reducing the temperature T
```{r}
#data
y <- c(0.9411, 1.9510, 3.1513)
dy <- c(0.11, 0.10, 0.13)
t <- c(1:length(y))

#chi^2-function
getX2 <- function(param, y, dy, t){
  chi2 <- sum(((param[1]*t + param[2]) - y)**2/dy**2)
  return(chi2)
}

#simulated annealing
simann <- function(y, dy, t, Temp, alpha){
  steps <- 100
  stepsTemp <- 100
  xi <- c(1, 0)
  chi2old <- getX2(xi, y, dy, t)
  minvals <- xi
  while(steps > 0){
    while(stepsTemp > 0){
      xi[1] <- rnorm(n=1, mean = 1, sd = 1)
      xi[2] <- rnorm(n=1, mean = 0, sd = 1)
      chi2new <- getX2(xi, y, dy, t)
      accept <- FALSE
      if(chi2new < chi2old) accept <- TRUE
      else if(runif(1) < exp((chi2old-chi2new)/Temp)) accept <- TRUE
      if(accept){
        minvals <- xi
        chi2old <- chi2new
      }
      stepsTemp <- stepsTemp - 1
    }
    stepsTemp = 100
    Temp <- alpha * Temp
    steps <- steps - 1
  }
  return(minvals)
}

#optimal values for X^2 and parameter a,b:
param <- c(1,0)
myfit <- optim(par = param, fn=getX2, y=y, dy=dy, t=t)
print("With R calculated X^2:")
print(myfit$value)
print("With R calculated parameters a, b:")
print(myfit$par)

#minimal X^2 and parameter a,b calculated with simulated annealing:
minvals <- simann(y,dy,t ,100, 0.9)
print("Simulated Annealing:")
print("Parameters a, b:")
print(minvals)
print("X^2:")
print(getX2(minvals,y,dy,t))

plot(x=t, y=y, ylim=c(0.5,3.5), col = "blue", main = "y(t) with fit ")
arrows(x0=t, y0=y-dy, x1=t, y1=y+dy, length= 0.05, angle=90, code=3, col = "blue")
lines(x=t, y=(myfit$par[1]*t + myfit$par[2]), col = "red")
lines(x=t, y=(minvals[1]*t + minvals[2]), col = "green")

```